Disentangling linkages between satellite derived forest structure and productivity essential biodiversity variables.

### 2023-06-06 scripts/manuscript/manuscript.qmd
send to all coauthors

abstract: Under a changing climate and in light of the biodiversity crises, monitoring change in biodiversity at broad scales has become ever more imperative. Key to regular biodiversity monitoring over time and space is the definition and selection of variables capable of accurately monitoring biodiversity trends. One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables, and capable of monitoring biodiversity globally while also being complementary to one another, within EBV classes. Five of six EBV classes have been shown to be well suited to be monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes. In this paper, we seek to address the complementarity of variables describing forest structure, as imputed across the entirety of British Columbia, Canada, with variables representing ecosystem functioning, namely the Dynamic Habitat Indices, a yearly summary of productivity indices. Both sets of variables have previously been linked with biodiversity metrics across a range of scales. Using redundancy analysis, we find that forest structural attributes and the DHIs are essentially decoupled from each other, with the forest structure variables explaining 14% of the variation in the DHIs. Further, we explore how the proportion of variance explained in the DHIs varies with either primary structure variables, such as canopy height and cover, vs more complex modelled forest structural attributes, such as aboveground biomass and volume. Results suggest there is large overlap between the two types of forest structure variables suggesting little added benefit of derived attributes beyond height and cover, in this context. Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV represent markedly different processes as observed by remote sensing and they are highly complementary across large environmental gradients, suggesting that in an earth-observation based monitoring program both are needed.

@@

With biodiversity being in decline, and facing extinction rates above
the background extinction rate [@thomas2004; @urban2015], as well as the
homogenization of communities at various scales [@mcgill2015] it is
integral to be able to monitor how biodiversity is changing across the
globe. In response, the global biodiversity community is making efforts
to assess and halt the degradation of biodiversity. The Group for Earth
Observation Biodiversity Observation Network has developed the Essential
Biodiversity Variables [EBVs, @pereira2013], designed as an analog to
the Essential Climate Variables framework [@bojinski2014]. EBVs are
designed to be global in scope, relevant to biodiversity information,
feasible to use, and complementary to one another [@skidmore2021]. While
it can be difficult, time consuming, and expensive to collect data on
biodiversity across wide swaths of land and varying ecosystems, EBVs,
which can be correlated to sampled biodiversity information, allow for
the monitoring and assessment of protected area effectiveness and
ecosystem health at large spatial scales [@hansen2021]. There are six
EBV classes, each of which correspond to a different facet of
biodiversity, including species populations, species traits, community
composition, ecosystem structure, ecosystem function, and genetic
composition [@pereira2013].

Satellite remote sensing has proven to be capable of measuring five of
the six EBV classes, the exception being genetic composition, which
requires in-situ observation and sampling [@skidmore2021]. Species
populations - and in turn community composition - can be assessed with
very-high-resolution imagery to identify tree species at the tree-crown
scale, however it is difficult and computationally expensive to extend
these analyses to broader extents [@fassnacht2016; @graves2016] while
species traits such as vegetation phenology have been observed at the
single-tree scale using, for example, PlanetScope imagery and
drone-based measurements [@wu2021]. However, the spatially limited and
often *ad-hoc* data collection approaches associated with monitoring
individuals is not conducive to the global or regional scales required
for biodiversity trend assessment [@valdez2023].

The two landscape-level EBVs (ecosystem structure and function) are well
suited to be examined at large spatial scales using coarser spatial
measurements, such as those taken from satellites by the Moderate
Resolution Imaging Spectroradiometer [MODIS, @zhang2003], the Landsat
imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
@darvishzadeh2019] programs. These mid-resolution satellites can monitor
processes at broader extents but the coarse spatial resolution removes
the ability to relate these traits to individual organisms. As a result,
satellite remote sensing data has been shown to be arguably the most
effective at monitoring ecosystem based EBVs focused on structure and
function. These EBVs classes can be monitored at regional to global
extents through the use of optical imagery [@cohen2004], as well as
active sensors such as lidar (light detection and ranging) and radar
[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].

@@

Increased forest structural complexity has been hypothesized to create
additional niches, leading to increased species diversity [@bergen2009],
which has been frequently demonstrated using avian species diversity
metrics [@macarthur1961]. For example: @herniman2020 used spectral and
lidar derived forest structure data to model avian habitat suitability;
@clawges2008 found that lidar derived forest structural attributes are
capable of identifying habitat types associated with avian species in
pine/aspen forests; @goetz2007 used canopy structural diversity to
predict bird species richness, finding that canopy vertical distribution
was the strongest predictor of species richness. Forest structural
metrics have also been used to study biodiversity in other clades as
well [@davies2014; @nelson2005].

@@

@berry2007 first explored this idea by proposing the linkage of
intra-annual summaries of MODIS derived GPP to dispersive bird species.
This idea was further refined into the Dynamic Habitat Indices [DHIs,
@coops2008], which have now been shown to be well suited to assess the
three aforementioned hypotheses at global scales [@radeloff2019]. The
cumulative DHI calculates the total amount of energy available in a
given pixel over the course of a year. Cumulative DHI is strongly linked
to the available energy hypothesis, which suggests that with greater
available energy species richness will increase [@wright1983]. The
minimum DHI, which calculates the lowest productivity over the course of
a year can be matched to the environmental stress hypothesis, which
proposes that higher levels of minimum available energy will lead to
higher species richness [@currie2004]. Finally, the variation DHI, which
calculates the coefficient of variance in a vegetation index through the
course of a year, corresponds to the environmental stability hypothesis
which states that lower energy variation throughout a year will lead to
increased species richness [@williams2008].

@@

Linkages between forest ecosystem structure and function have been
examined within a remote sensing context for over 20 years [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship at single time points
(within a single image) [@myneni1994], various relationship directions
and shapes have been found between forest structure and function metrics
[@ali2019]. Hypothesized mechanisms such as niche complementary have
shown that aboveground biomass increases with stand structure
[@zhang2012], while asymmetric competition for light can reduce forest
productivity with increased structural complexity [@bourdier2016]. The
relationship in particular between forest structural diversity metrics -
which are now more accurately and comprehensively derived from lidar
data - and temporal variation in functional metrics, specifically the
metrics of ecosystem productivity via the DHI framework, have yet to be
fully examined.

The overall goal of this paper is to assess patterns of forest ecosystem
structure and function, as observed using remote sensing technologies,
and their complementarity across a wide range of ecosystems encompassing
significant environmental gradients. To do so, we synthesize data from
moderate-scale remote-sensing derived metrics of forest structure,
represented as both simple ALS-extracted metrics of canopy height, cover
and vertical complexity, as well as modelled forest structure attributes
including volume, and aboveground biomass, with a well-established
remote sensing derived index on ecosystem function. First, we examine
how ecosystem structure and function complement one another across a
large environmental gradient and then compare the simple and modelled
representations of forest structure to different levels of ecosystem
function. This question is important as it provides insights to the EBV
community around complementarity of remote sensing metrics when
describing the structure and function of ecosystems and proposes a
method to examine potential overlap when generating remote sensing EBVs.

Our second question examines the independent and shared relationships of
ecosystem structure height and cover, with modelled forest structure, on
ecosystem function. This provides insight into the choice of remote
sensing attributes to use when developing EBVs within a single EBV
class. Remote sensing datasets can comprise relatively unprocessed
observations, in this case ALS measures of height and cover which are
derived from the raw 3D point cloud vs modelled attributes, such as
biomass and volume, which involve the use the statistical relationships
with field data to transform the observations into more refined data
products. Assessing which of these two (or combination of the two)
approaches has stronger or weaker relationships with estimates of
ecosystem function provides insights into the choice of data used to
build EBVs. Lastly, we examine how the primary and modelled structure
variables partition the variance of the DHIs within key biomes and
forest types across a large environmental range, examining to what
extent ecosystem and forest types impacts these relationships and thus
providing insight into the applicability of these results globally.

@@

These zones can be further split into subzones, variants, and phases
based on microclimate, precipitation, and topography [@pojar1987]. To
examine trends across the large environmental gradients, we group the
BEC zones into five broad biomes, specifically, the southern interior,
northern interior, montane, alpine, and coastal groups similar to
@hamann2006. We also report each BEC zone's average climate data,
according to @wang2020 (@tbl-bec-group).

@@

We use an established set of indices of annual productivity shown to be
related to global biodiversity trends: the Dynamic Habitat Indices
[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
productivity variables that summarize the cumulative amount of available
energy, the minimum available energy, and the variation in available
energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
have previously been produced at a global extent using MODIS imagery,
and have previously been used to assess alpha [@radeloff2019] and beta
[@andrew2012] diversity, species abundances [@razenkova2023], and
construct novel ecoregionalizations [@coops2009; @andrew2013] . Recent
studies have began to examine how these indices can be constructed at a
finer spatial resolution by using multi-annual Landsat imagery to
generate a single synthetic year of monthly observations
[@razenkova2022].

>>>>

abstract: Under a changing climate and in light of the biodiversity crises, monitoring change in biodiversity at broad scales has become ever more imperative. Key to regular biodiversity monitoring over time and space is the definition and selection of variables capable of accurately monitoring biodiversity trends. One framework is the Essential Biodiversity Variables (EBV), which are designed to be analogous to the Essential Climate variables and capable of monitoring biodiversity globally while also being complementary to one another within EBV classes. Five of six EBV classes are well suited to being monitored using satellite remote sensing, with particular focus warranted on the ecosystem structure and function classes. In this paper, we seek to address the complementarity of variables describing forest structure, as imputed across the entirety of British Columbia, Canada, with variables representing ecosystem functioning, namely the Dynamic Habitat Indices, a yearly summary of productivity indices. Both sets of variables have previously been linked with biodiversity metrics across a range of scales. Using redundancy analysis, we find that forest structural attributes and the DHIs are essentially decoupled from each other, with the forest structure variables explaining 14% of the variation in the DHIs. Further, we explore how the proportion of variance explained in the DHIs varies with either primary structure variables, such as canopy height and cover, vs more complex modelled forest structural attributes, such as aboveground biomass and volume. Results suggest there is large overlap between the two types of forest structure variables suggesting little added benefit of derived attributes beyond height and cover, in this context. Overall, we find that forest structure as an ecosystem structure EBV and the DHIs as an ecosystem function EBV represent markedly different processes as observed by remote sensing and they are highly complementary across large environmental gradients, suggesting that in an earth-observation based monitoring program both are needed.

@@

With biodiversity in decline, and facing extinction rates above the
background extinction rate [@thomas2004; @urban2015], as well as the
homogenization of communities at various scales [@mcgill2015], it is
integral to be able to monitor how biodiversity is changing across the
globe. In response, the global biodiversity community is making efforts
to assess and halt the degradation of biodiversity. The Group on Earth
Observations Biodiversity Observation Network has developed the
Essential Biodiversity Variables [EBVs, @pereira2013], designed as an
analog to the Essential Climate Variables framework [@bojinski2014].
EBVs are designed to be global in scope, relevant to biodiversity
information, feasible to use, and complementary to one another
[@skidmore2021]. While it can be difficult, time consuming, and
expensive to collect data on biodiversity across wide swaths of land and
varying ecosystems, EBVs, which can be correlated to sampled
biodiversity information, allow for the monitoring and assessment of
protected area effectiveness and ecosystem health at large spatial
scales [@hansen2021]. There are six EBV classes, each of which
correspond to a different facet of biodiversity: species populations,
species traits, community composition, ecosystem structure, ecosystem
function, and genetic composition [@pereira2013].

Satellite remote sensing has proven to be capable of measuring five of
the six EBV classes, the exception being genetic composition, which
requires in-situ observation and sampling [@skidmore2021]. Species
populations, and in turn community composition, can be assessed with
very-high-resolution imagery to identify tree species at the tree-crown
scale [@fassnacht2016; @graves2016]. Additionally, species traits such
as vegetation phenology have been observed at the single-tree scale
using, for example, PlanetScope imagery and drone-based measurements
[@wu2021]. However, the spatially limited and often *ad-hoc* data
collection approaches associated with monitoring individuals is not
conducive to the global or regional scales required for biodiversity
trend assessment [@valdez2023], in addition to the difficulties
associated with extending species-level analyses to broader extents
[@fassnacht2016; @graves2016].

The two landscape-level EBVs (ecosystem structure and function) are well
suited to be examined at large spatial scales using coarser spatial
measurements, such as those taken from satellites by the Moderate
Resolution Imaging Spectroradiometer [MODIS, @zhang2003], the Landsat
imaging systems [@fisher2006], or Sentinel-2 [@helfenstein2022;
@darvishzadeh2019] programs. These mid-resolution satellites can monitor
processes at broader extents, but their coarse spatial resolution
removes the ability to relate these traits to individual organisms. As a
result, satellite remote sensing data has been shown to be arguably the
most effective at monitoring ecosystem based EBVs focused on structure
and function. These EBVs classes can be monitored at regional to global
extents through the use of optical imagery [@cohen2004], as well as
active sensors such as lidar (light detection and ranging) and radar
[@guo2017; @lefsky2002; @lang2021; @neuenschwander2019; @coops2016].

@@

Increased forest structural complexity has been hypothesized to create
additional niches, leading to increased species diversity [@bergen2009],
which has been frequently demonstrated using avian species diversity
metrics [@macarthur1961]. For example, @herniman2020 used spectral and
lidar derived forest structure data to model avian habitat suitability;
@clawges2008 found that lidar derived forest structural attributes are
capable of identifying habitat types associated with avian species in
pine/aspen forests; @goetz2007 used canopy structural diversity to
predict bird species richness, finding that canopy vertical distribution
was the strongest predictor of species richness. Forest structural
metrics have also been used to study biodiversity in other clades as
well [@davies2014; @nelson2005].

@@

@berry2007 first explored this idea by proposing the linkage of
intra-annual summaries of MODIS-derived GPP to dispersive bird species.
This idea was further refined into the Dynamic Habitat Indices [DHIs,
@coops2008], which have now been shown to be well suited to assess the
three aforementioned hypotheses at global scales [@radeloff2019]. The
cumulative DHI calculates the total amount of energy available in a
given pixel over the course of a year. Cumulative DHI is strongly linked
to the available energy hypothesis, which suggests that with greater
available energy species richness will increase [@wright1983]. The
minimum DHI, which calculates the lowest productivity over the course of
a year can be matched to the environmental stress hypothesis, which
proposes that higher levels of minimum available energy will lead to
higher species richness [@currie2004]. Finally, the variation DHI, which
calculates the coefficient of variance in a vegetation index through the
course of a year, corresponds to the environmental stability hypothesis
which states that lower energy variation throughout a year will lead to
increased species richness [@williams2008].

@@

Linkages between forest ecosystem structure and function have been
examined within a remote sensing context for over 20 years [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship at single time points
(e.g. within a single image) [@myneni1994], various relationship
directions and shapes have been found between forest structure and
function metrics [@ali2019]. Hypothesized mechanisms such as niche
complementarity have shown that aboveground biomass increases with stand
structure [@zhang2012], while asymmetric competition for light can
reduce forest productivity with increased structural complexity
[@bourdier2016]. The relationship in particular between forest
structural diversity metrics - which are now more accurately and
comprehensively derived from lidar data - and temporal variation in
functional metrics, specifically the metrics of ecosystem productivity
via the DHI framework, have yet to be fully examined.

The overall goal of this paper is to assess patterns of forest ecosystem
structure and function, as observed using remote sensing technologies,
and their complementarity across a wide range of ecosystems encompassing
significant environmental gradients. To do so, we synthesize data from
moderate-scale remote-sensing derived metrics of forest structure,
represented as both simple ALS-extracted metrics of canopy height, cover
and vertical complexity, as well as modelled forest structure attributes
including volume and aboveground biomass, with a well-established remote
sensing derived index on ecosystem function. First, we examine how
ecosystem structure and function complement one another across a large
environmental gradient and then compare the simple and modelled
representations of forest structure to different levels of ecosystem
function. This question is important as it provides insights to the EBV
community around complementarity of remote sensing metrics when
describing the structure and function of ecosystems and proposes a
method to examine potential overlap when generating remote sensing EBVs.

Our second question examines the independent and shared relationships of
lidar derived forest structural attributes, with modelled forest
structure, on ecosystem function. This provides insight into the
selection of remote sensing attributes to use when developing EBVs
within a single EBV class. Remote sensing datasets can comprise
relatively unprocessed observations, in this case lidar-imputed measures
of height and cover vs modelled attributes, such as biomass and volume,
which involve the use the statistical relationships with field data to
transform the observations into more refined data products. Assessing
which of these two (or combination of the two) approaches has stronger
or weaker relationships with estimates of ecosystem function provides
insights into the choice of data used to build EBVs. Lastly, we examine
how the primary and modelled structure variables partition the variance
of the DHIs within key biomes and forest types across a large
environmental range, examining to what extent ecosystem and forest types
impacts these relationships and thus providing insight into the
applicability of these results globally.

@@

These zones can be further split into subzones, variants, and phases
based on microclimate, precipitation, and topography [@pojar1987]. To
examine trends across the large environmental gradients, we group the
BEC zones into five broad biomes: the southern interior, northern
interior, montane, alpine, and coastal groups similar to @hamann2006. We
also report each BEC zone's average climate data, according to @wang2020
(@tbl-bec-group).

@@

We use an established set of indices of annual productivity shown to be
related to global biodiversity trends: the Dynamic Habitat Indices
[@radeloff2019]. The DHIs are a set of satellite remote sensing derived
productivity variables that summarize the cumulative amount of available
energy, the minimum available energy, and the variation in available
energy throughout a given year [@berry2007; @radeloff2019]. The DHIs
have previously been produced at a global extent using MODIS imagery,
and have previously been used to assess alpha [@radeloff2019] and beta
[@andrew2012] diversity, species abundances [@razenkova2023], and
construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
studies have began to examine how these indices can be constructed at a
finer spatial resolution by using multi-annual Landsat imagery to
generate a single synthetic year of monthly observations
[@razenkova2022].
